Tight battle between republicans and democrats with Kamala in the lead

### 2024-11-04 paper/paper.qmd
Update paper.qmd
Added model, results and discussion, and fixed formatting

library(tidyverse)
library(knitr)
library(kableExtra)
data <- read_csv("../data/02-analysis_data/analysis_data.csv")

just_harris_high_quality <- read_csv("../data/02-analysis_data/analysis_data_harris.csv")

just_trump_high_quality <- read_csv("../data/02-analysis_data/analysis_data_trump.csv")
```

@@

**Data Preparation**

@@

**Feature Selection**

@@

**Model Training and Prediction**

@@

**Model Assumptions**

@@

**Cross-Validation and Overfitting Prevention**

@@

**Why Multiple Linear Regression?**

@@

**Why These Specific Predictors?**

@@

**Justification for Model Complexity**

@@

**Limitations of the Models**

@@

**Alternatives Considered**

@@

**Evaluation Metrics**

@@

**Summary**

@@

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

library(tidyverse)
library(knitr)
library(kableExtra)
#data <- read_csv("../data/02-analysis_data/analysis_data.csv")
just_harris_high_quality <- read_csv("../data/02-analysis_data/analysis_data_harris.csv")
just_trump_high_quality <- read_csv("../data/02-analysis_data/analysis_data_trump.csv")
```

@@

## Data Preparation

@@

## Feature Selection

@@

## Model Training and Prediction

@@

## Model Assumptions

@@

## Cross-Validation and Overfitting Prevention

@@

## Why Multiple Linear Regression?

@@

## Why These Specific Predictors?

@@

## Justification for Model Complexity

@@

## Limitations of the Models

@@

## Alternatives Considered

@@

## Evaluation Metrics

@@

## Summary

@@

### Model data ####
# Model 1 for Kamala: pct as a function of end_date, transparency_score and numeric_date
model_date <- lm(pct ~ end_date + transparency_score + numeric_grade, data = just_harris_high_quality)

# Model 2 for Kamala: pct as a function of end_date and pollster
model_date_pollster <- lm(pct ~ end_date + pollster, data = just_harris_high_quality)

# Augment data with model predictions
just_harris_high_quality <- just_harris_high_quality |>
  mutate(
    fitted_date = predict(model_date),
    fitted_date_pollster = predict(model_date_pollster)
  )

# Model 1 for Trump: pct as a function of end_date, transparency_score and numeric_date
model_date_trump <- lm(pct ~ end_date  + transparency_score + numeric_grade, data = just_trump_high_quality)

# Model 2 for Trump: pct as a function of end_date and pollster
model_date_pollster_trump <- lm(pct ~ end_date + pollster, data = just_trump_high_quality)

# Augment Trump data with model predictions
just_trump_high_quality <- just_trump_high_quality |>
  mutate(
    fitted_date = predict(model_date_trump),
    fitted_date_pollster = predict(model_date_pollster_trump)
  )

# Extract key metrics for Model 1 (Harris and Trump)
metrics_model_1 <- tibble(
  Candidate = c("Kamala Harris", "Donald Trump"),
  R_squared = c(summary(model_date)$r.squared, summary(model_date_trump)$r.squared),
  Adj_R_squared = c(summary(model_date)$adj.r.squared, summary(model_date_trump)$adj.r.squared),
  AIC = c(AIC(model_date), AIC(model_date_trump)),
  RMSE = c(sqrt(mean(residuals(model_date)^2)), sqrt(mean(residuals(model_date_trump)^2)))
)

# Create a table of metrics for Model 1
knitr::kable(metrics_model_1, caption = "Key Metrics for Harris's and Trump's Model 1")

```
## Key Metrics for Model 1

Table 5 presents the key metrics for Model 1, which served as a baseline analysis for polling percentages for Kamala Harris and Donald Trump. The R-squared values are relatively low for both candidates, at 0.07 for Harris and 0.14 for Trump, indicating that this simple model explained only a small proportion of the variance in polling percentages. The AIC values were 4926.19 and 5053.99 for Harris and Trump respectively, with similar RMSE values around 4, suggesting limited predictive power. This baseline model was used to establish a simple trend before moving to a more comprehensive model that accounts for additional variability.
```{r}
#| echo: false
#| eval: true
#| label: Summary Statistics for Donald Trump
#| tbl-cap: "Summary Statistics for Donald Trump"
#| warning: false
library(modelsummary)
library(knitr)
summary_trump_model_2 <- summary(model_date_pollster_trump)
significant_trump_model_2 <- as.data.frame(summary_trump_model_2$coefficients) %>%
  filter(`Pr(>|t|)` < 0.05)

# Create a trimmed-down summary table using kable
knitr::kable(significant_trump_model_2, caption = "Significant Predictors in Trump's Model 2")
```

```{r}
#| echo: false
#| eval: true
#| label: Summary Statistics for Kamala Harris
#| tbl-cap: "Summary Statistics for Kamala Harris"
#| warning: false
library(modelsummary)
library(knitr)
summary_harris_model_2 <- summary(model_date_pollster)
significant_harris_model_2 <- as.data.frame(summary_harris_model_2$coefficients) %>%
  filter(`Pr(>|t|)` < 0.05)

# Create a trimmed-down summary table using kable
knitr::kable(significant_harris_model_2, caption = "Significant Predictors in Harris's Model 2")
```

## Significant Predictors in Model 2

Table 6 and Table 7 present the summary statistics for Model 2 for Donald Trump and Kamala Harris respectively. The second model extends the baseline model by including pollster-specific effects, which significantly improves its predictive capabilities.

For Donald Trump, significant predictors include end_date (estimate = 0.0206, p < 0.001), which shows that his polling percentage tends to increase slightly over time. Specific pollsters like Ipsos (estimate = 5.39, p < 0.001) and MassINC Polling Group (estimate = 8.88, p < 0.001) show considerable effects, indicating variability in results due to the polling organization.

For Kamala Harris, end_date is also significant (estimate = 0.0129, p = 0.0015), suggesting a positive trend over time. Notable pollster effects include MassINC Polling Group (estimate = 7.70, p < 0.001) and University of Maryland/Washington Post (estimate = 14.62, p < 0.001), indicating significant variability in the polling results reported by different organizations.

```{r}
#| echo: false
#| eval: true
#| label: Polling Prediction Outcome
#| tbl-cap: "Polling prediction Outcome"
#| warning: false
# Plot model predictions for Trump
# Combine the datasets for Trump and Harris, keeping their respective predictions
combined_data_with_preds <- bind_rows(
  just_harris_high_quality |> mutate(candidate = "Kamala Harris"),
  just_trump_high_quality |> mutate(candidate = "Donald Trump")
)

# Plotting the model predictions for both Trump and Harris in one chart
ggplot(combined_data_with_preds, aes(x = end_date, y = pct)) +
  # Actual data points with transparency
  geom_point(aes(color = candidate), alpha = 0.5) +

  # Predicted lines for each candidate
  geom_line(data = combined_data_with_preds |> filter(candidate == "Kamala Harris"),
            aes(y = fitted_date_pollster, color = candidate), linetype = "solid", size = 1) +
  geom_line(data = combined_data_with_preds |> filter(candidate == "Donald Trump"),
            aes(y = fitted_date_pollster, color = candidate), linetype = "solid", size = 1) +

  # Set colors manually for better distinction
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +

  # Add labels and theme
  labs(
    x = "Polling End Date ",
    y = "Polling Percentage",
    title = "Polling Predictions for Kamala Harris and Donald Trump",
    color = "Candidate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Plotting the smoothed predictions for both Trump and Harris
ggplot(combined_data_with_preds, aes(x = end_date, y = pct)) +
  # Actual data points with transparency
  geom_point(aes(color = candidate), alpha = 0.3) +

  # Smoothed lines for each candidate
  geom_smooth(aes(color = candidate), method = "loess", span = 0.2, se = FALSE, size = 1.2) +

  # Set colors manually for better distinction
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +

  # Add labels and theme
  labs(
    x = "Polling End Date",
    y = "Polling Percentage",
    title = "Smoothed Polling Predictions for Kamala Harris and Donald Trump",
    color = "Candidate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Visual Representation of Model Predictions

Figure 1 and Figure 2 depict the predicted polling percentages for both candidates over time based on Model 2. The solid lines represent the smoothed predictions for Kamala Harris (blue) and Donald Trump (red). The visualizations reveal that both candidates maintain relatively stable trends, with some fluctuation, particularly influenced by pollsters and specific dates.

The smoothed polling predictions in Figure 2 indicate that Kamala Harris tends to have a slightly higher polling percentage compared to Donald Trump, with less fluctuation over the observed period. This suggests that her polling results are more consistent across different pollsters and dates.

## Summary

The results indicate that while both candidates’ polling percentages are influenced by factors such as pollster-specific effects and end date, Kamala Harris shows more consistent support over time. The extended Model 2 provided better explanatory power compared to the baseline Model 1, as reflected in the inclusion of pollster effects that help explain the observed variability in polling data.

@@

## Key Findings

This paper provides a forecast for the 2024 U.S. Presidential Election between Kamala Harris and Donald Trump, leveraging linear regression models applied to recent polling data. The analysis projects Kamala Harris as the leading candidate, with consistent support across various polls and polling organizations, suggesting her potential to win the upcoming election.

Model 1, which included basic predictors like end date, transparency score, and numeric grade, demonstrated limited explanatory power, with R-squared values of 0.07 for Harris and 0.14 for Trump. However, Model 2, which introduced pollster-specific effects, provided better insights into the variability in polling percentages.

The smoothed polling predictions (Figure 2) clearly showed that Harris is ahead of Trump throughout the analyzed period. This consistent lead, combined with higher polling percentages reported by major polling organizations such as MassINC Polling Group and University of Maryland/Washington Post, indicates Harris’s current advantage heading toward the election.

## Pollster Impact and Voter Sentiment

Pollster-specific effects played a substantial role in the reported polling outcomes. For Kamala Harris, significant pollster effects included organizations like MassINC Polling Group and University of Maryland/Washington Post, which reported notably higher polling percentages for Harris. This suggests that certain polling organizations may have a favorable sampling bias or different methodologies that influence their reported results.

Similarly, Donald Trump also saw significant variability across pollsters. Notably, Ipsos and MassINC Polling Group were among those reporting significantly higher polling percentages for Trump. These pollster-specific biases suggest the importance of accounting for the polling organization when analyzing election forecasts.

## Regional Analysis and Swing States

The regional analysis highlighted in swing states shows varying degrees of support for both candidates, with some states showing Harris in the lead and others favoring Trump. Notable swing states like Arizona and Pennsylvania displayed significant variation, with Harris maintaining a narrow lead in some polls while Trump showed higher support in others.

The combined analysis of both Harris and Trump across the swing states suggests that Harris is predicted to outperform Trump, albeit with narrow margins in key regions. This is critical for electoral success, as even slight changes in voter sentiment in swing states can have a substantial impact on the final election outcome. The projections emphasize the importance of strategic campaigning in these battleground states, where the race remains close.

## Polling and Model Uncertainty

The polling data used in this analysis, while indicative of current trends, introduces uncertainties due to incomplete coverage and pollster bias. The variability observed across different pollsters underscores the challenges in capturing a fully representative sample of the voting population. Certain demographic groups may be underrepresented, resulting in biased outcomes.

The models assumed linear relationships between polling percentages and predictors, which may have oversimplified the complex relationships at play in voter sentiment. This linearity assumption, while useful for initial analysis, may overlook potential non-linear dynamics that can significantly affect polling outcomes, especially during dynamic campaign events or sudden shifts in public opinion.

## Limitations and Weaknesses

Despite the predictive capabilities of the models, several limitations should be acknowledged:

1.Assumption of Linearity: The assumption of linearity restricts the model’s ability to capture non-linear shifts in voter sentiment, particularly during pivotal campaign moments. Incorporating non-linear modeling techniques could better reflect the complexity of voter behavior.

2.Pollster Variability: The differences in poll results across organizations, as observed in Model 2, point to systematic biases in polling methodologies. These biases may reflect differences in sampling techniques, question wording, or demographic targeting, which can skew reported outcomes.

3.Potential Polling Biases: Nonresponse bias remains a significant challenge, particularly for Trump’s support, as seen in previous election cycles. The “Shy Tory Effect” may lead to an underreporting of support for Trump, making it difficult to capture the full extent of his voter base.

4.Historical Data and Recency Bias: The reliance on historical data risks the model being influenced by past trends that may not hold in the current electoral cycle. Unexpected shifts in voter sentiment, as seen in previous elections, could lead to discrepancies between the model predictions and actual outcomes.

## Implications for Future Research

The results of this analysis highlight areas for improvement and further research:

- Non-Linear Modeling: Future research could explore the use of non-linear models or splines to capture the complexity of voter sentiment, particularly during pivotal campaign events.
- Bayesian Methods for Dynamic Updates: Implementing a Bayesian framework would allow the model to update its predictions in real time as new polling data becomes available. This could help better capture the evolving nature of voter preferences, especially in response to major campaign developments.
- Incorporation of Demographic Data: Including demographic information—such as age, gender, income, and education—could provide more granular insights into voter behavior and help identify which demographic groups are most influential in determining electoral outcomes.
- Weighting Polls by Credibility: Incorporating a weighting scheme based on pollster credibility and transparency scores could help adjust for systematic biases and provide a more balanced view of the polling landscape.

## Conclusion

The analysis presented in this paper indicates a clear lead for Kamala Harris over Donald Trump in the 2024 U.S. Presidential Election, based on recent polling data. While the models capture broad trends in support and demonstrate the impact of polling methodology, they also emphasize the inherent uncertainty in electoral forecasting. The current projections suggest that Harris is positioned well to win, particularly if the trends observed in polling data continue. However, these results should be interpreted cautiously, acknowledging the dynamic nature of elections and the various limitations inherent in polling-based modeling.

Future studies should seek to incorporate additional data sources and more sophisticated modeling techniques to better capture the complex dynamics of voter behavior, particularly in an ever-evolving political environment.

@@

#plot(first_model, "trace")

#plot(first_model, "rhat")
```
